---
title: Quality assessment of cancer data
author:
- name: Berta Canal Simón
  affiliation:
  - &id Barcelona School of Economics
  email: berta.canal@bse.eu
- name: Adam Olivares Canal
  affiliation: *id
  email: adam.olivares@bse.eu
date: "`r format(Sys.time(), '%B %e, %Y')`"
abstract: >
  Here we perform a first exploratory analysis and quality assessment of the cfRNA expression profiles of the cancer data.
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    fig_captions: yes
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: _bibliography.bib
---

```{r setup, echo=FALSE, cache=FALSE, message=FALSE}
library(knitr)
library(kableExtra)
library(VIM)

knitr::opts_chunk$set(
  collapse=TRUE,
  comment="",
  fig.align="center",
  fig.wide=TRUE,
  cache=TRUE,
  cache.path="_cache/__QA_cn",
  cache.extra=R.version.string,
  autodep=TRUE
)

source(file.path("_scripts", "labelPanel.R")) ## for the labelPanel() function
source(file.path("_scripts", "myplotbcv.R")) ## for the myplotbcv() function
```

# Read molecular and phenotype data

Read the molecular and phenotype data stored in serialized `SummarizedExperiment`
objects.

```{r, message=FALSE}
library(SummarizedExperiment)

seR <- readRDS(file.path("_raw_data", "se.Roskams22.GENCODEv44.hg38.rds"))
dim(seR)
seB <- readRDS(file.path("_raw_data", "se.Block22.GENCODEv44.hg38.rds"))
dim(seB)
```

# Filtering of lowly-expressed genes and normalization

We will use the [edgeR](https://bioconductor.org/packages/edgeR) package for
processing the raw counts and doing some exploratory analyses. We start by
building the corresponding `DGEList` objects with the corresponding row and
column data.

```{r, message = FALSE, warning = FALSE}
library(edgeR)

dgeR <- DGEList(counts=assay(seR), genes=rowData(seR), samples=colData(seR))
dgeB <- DGEList(counts=assay(seB), genes=rowData(seB), samples=colData(seB))
```

Filter lowly expressed genes using the function `filterByExpr()` and the
`Tumor` phenotype column as grouping factor. We need to filter both,
the `SummarizedExperiment` and `DGEList` types of objects, to keep them
parallel to each other.

In the article, they sequenced plasma samples of eight hepatocellular carcinoma (HCC) and ten multiple myeloma (MM) patients, 12 patients of their respective precancerous conditions, and 20 non-cancer (NC) donors. For convenience, in the project will discern between those patients with and without cancer (yes and no levels), disregarding the type of tumor.

```{r}
seR@colData@listData[["Tumor"]] <- factor(seR$Disease != "NC", labels = c("no", "yes"))
dgeR$samples$Tumor <- factor(seR$Disease != "NC", labels = c("no", "yes"))

seB@colData@listData[["Tumor"]] <- as.factor(seB$Tumor)
dgeB$samples$Tumor <- as.factor(dgeB$samples$Tumor)

mask <- filterByExpr(dgeR, group=seR$Tumor)
sum(mask)
seR.filt <- seR[mask, ]
dim(seR.filt)
dgeR.filt <- dgeR[mask, , keep.lib.sizes=FALSE]
dim(dgeR.filt)

mask <- filterByExpr(dgeB, group=seB$Tumor)
sum(mask)
seB.filt <- seB[mask, ]
dim(seB.filt)
dgeB.filt <- dgeB[mask, , keep.lib.sizes=FALSE]
dim(dgeB.filt)
```

Calculate normalization factors on the filtered data.

```{r}
dgeR.filt <- calcNormFactors(dgeR.filt)
assays(seR.filt, withDimnames=FALSE)$logCPM <- cpm(dgeR.filt, log=TRUE)

dgeB.filt <- calcNormFactors(dgeB.filt)
assays(seB.filt)$logCPM <- cpm(dgeB.filt, log=TRUE)
```

Fo simplicity, change the name of the data from articles [@Roskams2022] and [PLACEHOLDER] to _training_ and _testing_, respectively.
```{r}
seR.filt.training <- seR.filt
dgeR.filt.training <- dgeR.filt

seB.filt.testing <- seB.filt
dgeB.filt.testing <- dgeB.filt
```

Figure \@ref(fig:mdsplot) below shows samples dissimilarity labeled by
the diagnosis of cancer in each of the two datasets.

```{r mdsplot, echo=FALSE, fig.height=4, fig.width=12, out.width="800px", fig.cap="Sample dissimilarity labeled by the diagnosis of cancer. Multidimensional scaling (MDS) plot of sample dissimilarity for the training (a) and testing (b) data from the study by @Roskams2022, and for the data from the study by [PLACEHOLDER]."}
tumorcolors <- c("skyblue", "darkred")
par(mfrow=c(1, 2), mar=c(4, 5, 2, 1))
xx <- plotMDS(dgeR.filt.training, labels=rep("", ncol(seR.filt.training)), las=1)
points(xx$x, xx$y, col=tumorcolors[as.integer(seR.filt.training$Tumor)],
       pch=19, cex=0.8)
legend("topleft", c("Control", " Cancer"), fill=tumorcolors,
       inset=0.01)
labelPanel("a")
xx <- plotMDS(dgeB.filt.testing, labels=rep("", ncol(seB.filt.testing)), las=1)
points(xx$x, xx$y, col=tumorcolors[as.integer(seB.filt.testing$Tumor)],
       pch=19, cex=0.8)
legend("topleft", c("Control", "Cancer"), fill=tumorcolors,
       inset=0.01)
labelPanel("b")
```

Save `DGEList` and `SummarizedExperiment` objects including filtered data,
normalization factors and normalized log CPM units of expression.

# Data imputation (PROVISIONAL)

## Testing dataset

Impute missing Age values by mean.

```{r}
mean.ageB <- mean(dgeB.filt.testing$samples$Age, na.rm = TRUE)
is.na.ageB <- which(is.na(dgeB.filt.testing$samples$Age) == TRUE)
dgeB.filt.testing$samples$Age[is.na.ageB] <- mean.ageB
seB.filt.testing$Age[is.na.ageB] <- mean.ageB
```

The estimation of missing data for `Sex` covariate uses a KNN imputer (k = 5) based on the counts for Y-linked genes (genes located on the Y chromosome). Although no Y-linked genes counts should be present in female samples, it does not hold for all samples in our dataset (maybe due to technical reasons). This led to disregarding the criteria of imputing based on the assignment of male gender to those individuals with count different than 0 for the Y-linked genes.

```{r}
mask <- grepl("Y-linked", dgeB.filt.testing$genes$Description)
counts_Y_linked <- dgeB.filt.testing$counts[mask,]
counts_Y_linked <- as.data.frame(t(counts_Y_linked))
counts_Y_linked$Sex <- as.factor(dgeB.filt.testing$samples$Sex)
data_imputed <- kNN(counts_Y_linked, k=5, variable="Sex")

dgeB.filt.testing$samples$Sex <- data_imputed$Sex
seB.filt.testing$Sex <- data_imputed$Sex
```

```{r}
saveRDS(dgeR.filt, file=file.path("_processed_data", "dgeR.filt.rds"))
saveRDS(seR.filt, file=file.path("_processed_data", "seR.filt.rds"))

saveRDS(dgeB.filt, file=file.path("_processed_data", "dgeB.filt.rds"))
saveRDS(seB.filt, file=file.path("_processed_data", "seB.filt.rds"))

saveRDS(dgeR.filt.training,
        file=file.path("_processed_data", "dgeR.filt.training.rds"))
saveRDS(seR.filt.training,
        file=file.path("_processed_data", "seR.filt.training.rds"))
saveRDS(dgeB.filt.testing,
        file=file.path("_processed_data", "dgeB.filt.testing.rds"))
saveRDS(seB.filt.testing,
        file=file.path("_processed_data", "seB.filt.testing.rds"))
```

# Dispersion estimates and hypervariable genes

We calculate dispersion estimates and biological coefficients of variation
[@mccarthy2012differential] and use them to identify hypervariable genes for
different models of increasing complexity that will help us to select the most
sensible one for the differential expression analysis between
Cancer-affected and unafected individuals.

The four considered models are: (1) a model with the intercept term only,
(2) a model with `Tumor` as main explanatory variable, (3) a model
with `Cancer` as main explanatory variable and `Age` and `Sex` as
covariates, and (4) a model with `Tumor` as main explanatory variable,
and `Age`, `Sex` and surrogate variables (SVs), calculated with SVA
[@leek2007capturing] as covariates. The estimation of SVs will be done using
only the expression values of the top-10% most variable genes.

## Training dataset

```{r, message=FALSE}
library(sva)

mod <- model.matrix(~ 1, colData(seR.filt.training))
dgeR.filt.training.interceptonly <- estimateDisp(dgeR.filt.training, design=mod,
                                                 robust=TRUE)
summary(dgeR.filt.training.interceptonly$prior.df)

mod <- model.matrix(~ Tumor, colData(seR.filt.training))
dgeR.filt.training.PE <- estimateDisp(dgeR.filt.training, design=mod,
                                      robust=TRUE)
summary(dgeR.filt.training.PE$prior.df)

mod <- model.matrix(~ Tumor + Age + Sex, colData(seR.filt.training))
dgeR.filt.training.PEGA <- estimateDisp(dgeR.filt.training, design=mod,
                                        robust=TRUE)
summary(dgeR.filt.training.PEGA$prior.df)

mod0 <- model.matrix(~ Age + Sex, colData(seR.filt.training))
IQRs <- apply(assays(seR.filt.training)$logCPM, 1, IQR)
mask <- IQRs > quantile(IQRs, prob=0.9)
sv <- sva(assays(seR.filt.training)$logCPM[mask, ],
          mod=mod, mod0=mod0)
mod <- cbind(mod, sv$sv)
dgeR.filt.training.PEGASVs <- estimateDisp(dgeR.filt.training,
                                           design=mod, robust=TRUE)
summary(dgeR.filt.training.PEGASVs$prior.df)
```

Figure \@ref(fig:Mtrainingavglogcpmxvbcv) shows, for each considered model, the
biological coefficient of variation (BCV) as function of the average expression,
in log CPM units.

```{r Mtrainingavglogcpmxvbcv, echo=FALSE, message = FALSE, fig.height=8, fig.width=8, dpi=100, results="hide", fig.cap="Biological coefficient of variation (BCV). BCV values as function of the average gene expression in log-CPM units."}
rngbcv <- range(sqrt(c(getDispersion(dgeR.filt.training.interceptonly),
                       getDispersion(dgeR.filt.training.PE),
                       getDispersion(dgeR.filt.training.PEGA),
                       getDispersion(dgeR.filt.training.PEGASVs))))
rngbcv[2] <- ceiling(rngbcv[2])
par(mfrow=c(2, 2), mar=c(4, 5, 2, 1))
syms <- mcols(seR.filt.training)$Symbol
myplotbcv(dgeR.filt.training.interceptonly, "~ 1", rngbcv, syms, cutoffdisp1=1.5,
          cutoffdisp2=0.8, cutoffA=9)
myplotbcv(dgeR.filt.training.PE, "~ Tumor", rngbcv, syms, cutoffdisp1=1.4,
          cutoffdisp2=0.8, cutoffA=9)
myplotbcv(dgeR.filt.training.PEGA, "~ Tumor + Age + Sex", rngbcv, syms,
          cutoffdisp1=1.4, cutoffdisp2=0.8, cutoffA=9)
myplotbcv(dgeR.filt.training.PEGASVs, "~ Tumor + Age + Sex + SVs",
          rngbcv, syms, cutoffdisp1=0.9, cutoffdisp2=0.8, cutoffA=9)
```

Table \@ref(tab:hypervargenesMtrainingcn) below shows the top-20 hypervariable genes
with lowest prior degrees of freedom for the last model.

```{r hypervargenesMtrainingcn, echo=FALSE}
ord <- order(dgeR.filt.training.PEGASVs$prior.df)
dtf <- data.frame(prior.df=dgeR.filt.training.PEGASVs$prior.df[ord],
                  BCV=sqrt(dgeR.filt.training.PEGASVs$tagwise.dispersion[ord]),
                  Gene=dgeR.filt.training.PEGASVs$genes$Symbol[ord],
                  Description=dgeR.filt.training.PEGASVs$genes$Description[ord])
ktab <- kable(dtf[1:20, ], "html", escape=FALSE, row.names=FALSE,
              caption="Top-20 hypervariable genes for the model Tumor + Sex + Age + SVs.")
ktab <- kable_styling(ktab, bootstrap_options=c("striped", "hover", "responsive"),
                      fixed_thead=TRUE)
kable_styling(ktab, position="center")
```

## Testing data

```{r, message=FALSE}
mod <- model.matrix(~ 1, colData(seB.filt.testing))
dgeB.filt.testing.interceptonly <- estimateDisp(dgeB.filt.testing, design=mod,
                                                robust=TRUE)
summary(dgeB.filt.testing.interceptonly$prior.df)
mod <- model.matrix(~ Tumor, colData(seB.filt.testing))
dgeB.filt.testing.PE <- estimateDisp(dgeB.filt.testing, design=mod,
                                     robust=TRUE)
summary(dgeB.filt.testing.PE$prior.df)
mod <- model.matrix(~ Tumor + Age + Sex, colData(seB.filt.testing))
dgeB.filt.testing.PEGA <- estimateDisp(dgeB.filt.testing, design=mod,
                                       robust=TRUE)
summary(dgeB.filt.testing.PEGA$prior.df)
mod0 <- model.matrix(~ Age + Sex, colData(seB.filt.testing))
IQRs <- apply(assays(seB.filt.testing)$logCPM, 1, IQR)
mask <- IQRs > quantile(IQRs, prob=0.9)
sv <- sva(assays(seB.filt.testing)$logCPM[mask, ],
          mod=mod, mod0=mod0)
mod <- cbind(mod, sv$sv)
dgeB.filt.testing.PEGASVs <- estimateDisp(dgeB.filt.testing,
                                          design=mod, robust=TRUE)
summary(dgeB.filt.testing.PEGASVs$prior.df)
```

```{r Mtestingavglogcpmxvbcv, echo=FALSE, fig.height=8, fig.width=8, dpi=100, results="hide", fig.cap="Biological coefficient of variation (BCV). BCV values as function of the average gene expression in log-CPM units."}
rngbcv <- range(sqrt(c(getDispersion(dgeB.filt.testing.interceptonly),
                       getDispersion(dgeB.filt.testing.PE),
                       getDispersion(dgeB.filt.testing.PEGA),
                       getDispersion(dgeB.filt.testing.PEGASVs))))
rngbcv[2] <- ceiling(rngbcv[2])
par(mfrow=c(2, 2), mar=c(4, 5, 2, 1))
syms <- mcols(seB.filt.testing)$Symbol
myplotbcv(dgeB.filt.testing.interceptonly, "~ 1", rngbcv, syms, cutoffdisp1=10.4,
          cutoffdisp2=0.8, cutoffA=9)
myplotbcv(dgeB.filt.testing.PE, "~ Tumor", rngbcv, syms, cutoffdisp1=10.4,
          cutoffdisp2=0.8, cutoffA=9)
myplotbcv(dgeB.filt.testing.PEGA, "~ Tumor + Age + Sex", rngbcv, syms,
          cutoffdisp1=10.4, cutoffdisp2=0.8, cutoffA=9)
myplotbcv(dgeB.filt.testing.PEGASVs, "~ Tumor + Age + Sex + SVs",
          rngbcv, syms, cutoffdisp1=10.4, cutoffdisp2=0.6, cutoffA=8)
```

```{r hypervargenesDsubset, echo=FALSE}
ord <- order(dgeB.filt.testing.PEGASVs$prior.df)
dtf <- data.frame(prior.df=dgeB.filt.testing.PEGASVs$prior.df[ord],
                  BCV=sqrt(dgeB.filt.testing.PEGASVs$tagwise.dispersion[ord]),
                  Gene=dgeB.filt.testing.PEGASVs$genes$Symbol[ord],
                  Description=dgeB.filt.testing.PEGASVs$genes$Description[ord])
ktab <- kable(dtf[1:20, ], "html", escape=FALSE, row.names=FALSE,
              caption="Top-20 hypervariable genes for the model Tumor + Age + Sex + SVs.")
ktab <- kable_styling(ktab, bootstrap_options=c("striped", "hover", "responsive"),
                      fixed_thead=TRUE)
kable_styling(ktab, position="center")
```

# Session information

```{r, child="_session-info.Rmd"}
```

# References


